M\"obius Inversion: From Posets to Categories

### 2017-04-12 Drafts/Draft.tex
editing error bounding

In fact, we observe exactly that behavior. Figure \ref{fig:prod_mag} shows the magnitudes of the three spaces as we vary $t$. At every value of $t$, $\chi_t(M)$ is almost exactly the product of the magnitudes of the other two spaces at that $t$ value.
\begin{figure}[H]
\[
\makebox[\textwidth][c]{\includegraphics[width = 0.9\paperwidth]{../Figures/prod_mag.png}}
\]
\caption{The magnitude of $M$ alongside the magnitudes of the factor spaces against $\log(t)$.}
\label{fig:prod_mag}
\end{figure}
\end{examp}
\section{Approximations and Error Bounding}
In Example \ref{ex:4pt}, $M$ was \emph{almost} a product space, so we expected its magnitude to be very nearly the product of the magnitudes of the two 2-point spaces. It would be useful to quantify how similar the magnitudes of two spaces are, given that we know something about how similar those spaces are. To do that, we need some simple tools from linear algebra involving vector and matrix norms. Note that these norms are not directly related to the norms we defined on monoidal categories - both are non-negative real-valued functions, but they operate on different structures.

@@

\item
\[
||A + B|| \leq ||A|| + ||B||.
\]
\item 
\[
||A|| = 0 \iff A = 0.
\]
\end{enumerate}
\end{lemma}
\begin{proof}
\begin{description}
\item[Spectral Norm:]
\begin{enumerate}
\item By the definition of the spectral norm, for any $v \in \Rr^n$,
\begin{align*}
|ABv| &\leq |A(||B||_s \times |v|)|\\
&\leq (||A||_s \times ||B||_s)|v|.
\end{align*}
It follows that 
 \[
||AB|| \leq ||A||_s\times ||B||.
\]
\item For any $v \in \Rr^n$,
\begin{align*}
|(A + B)v| &= |Av + Bv|\\
 &\leq |Av| + |Bv|\\
 &\leq ||A||_s + ||B||_s)v.
\end{align*}
\item It is clear that if $A=0$ then $||A||_s=0$. Now suppose that $A$ is not zero. Then the column space of $A$ has nonzero dimension, so there exists a $v$ such that $Av \neq 0$. So $||A||_s \geq \frac{|Av|}{|v|} > 0$.
\end{enumerate}
\item[$L_1$ Norm:]
\begin{enumerate}
\item We have
\begin{align*}
|AB_ij| &= \left| \sum_{k = 1}^nA_{ik}B{kj}\right|\\
&\leq \sum_{k = 1}^n|A_{ik}||B{kj}|. \text{ Therefore}\\
||AB||_1 &\leq \sum_{i = 1}^n\sum_{j=1}^n\sum_{k=1}^n|A_{ik}||B{kj}|\\
&\leq \sum_{i = 1}^n\sum_{j=1}^n\sum_{k=1}^n\sum_{l=1}^n|A_{ik}||B_{lj}|\\
&= ||A||_1||B||.
\end{align*}
\item For each $i,j$, $|(A + B)_{ij}| = |A_{ij} + B_{ij}| \leq |A_{ij}| + |B_{ij}|$. The result follows.
\item If $A=0$ then $||A||_1=0$. On the other hand, if any entry of $A$ is nonzero, then $||A||_1 > 0$ by definition.
\end{enumerate}
\end{description}
\end{proof}

We would like to show some results about limits and series convergence of matrices. To do that, we need to establish some notation. 
\begin{defn}\label{def:mat_conv}
Let $A_i$ be an infinite sequence of $m \times n$ matrices, and let $||\cdot||$ be some matrix norm (the spectral norm, the $L_1$ norm, or another matrix norm). Suppose we have an $m \times n$ matrix $L$ such that 
\[
\lim_{i \rightarrow \infty} ||A_i - L|| = 0.
\]
Then we say that 
\[
\lim_{i \rightarrow \infty} A_i = L.
\]
\end{defn}
It turns out that if the matrices involved are finite-dimensional, the choice of norm does not matter - a sequence will either always converge or always diverge. So there is no need to specify the norm when stating a fact about matrix convergence.

@@

\begin{lemma}\label{lem:lim0}
Let $E$ be a real-valued $n \times n$ matrix.  Suppose that $||E||_s < 1$. Then:
\[
\lim_{i \rightarrow \infty}E^i = 0.
\]
\end{lemma}
\begin{proof}
Recall that the meaning of this statement is determined by Definition \ref{def:mat_conv}.

Let $||E|| < 1$. Let $\vec{0}$ represent be the zero matrix. We now use Lemma \ref{lem:norm_prop}:
\begin{align*}
\lim_{i \rightarrow \infty}|| E^i - \vec{0}|| &\leq \lim_{i \rightarrow \infty}|| E^i|| - ||\vec{0}|| \\
&\leq \lim_{i \rightarrow \infty}|| E||^i\\
&= 0.
\end{align*}
Then, by Definition \ref{def:mat_conv},
\[
\lim_{i \rightarrow \infty} E^i = 0.
\]
\end{proof}
We actually could have used any norm instead of the spectral norm. But the spectral norm is always less than or equal to the $L_1$ norm, as well as most other matrix norms, so the more general result is implied.

This allows us to prove a useful fact which allows us to approximate the inverses of certain matrices.
\begin{thm}\label{thm:approx}
Let $A$ be an invertible $n \times n$ matrix, and let $E$ be an $n \times n$ matrix such that $||E|| < \frac 1{||A^{-1}||}$. Then $(A-E)$ is invertible, and
\begin{align*}
(A - E)^{-1} &= A^{-1} + A^{-1}EA^{-1} + A^{-1}EA^{-1}EA^{-1} + \cdots\\
&= A^{-1} + A^{-1}\sum_{i = 1}^\infty (EA^{-1})^i.
\end{align*}
\end{thm}
\begin{proof}
We apply Lemma \ref{lem:lim0}
\begin{align*}
(A-E)  A^{-1}\sum_{i = 0}^\infty (EA^{-1})^i
&= \left(\sum_{i = 0}^\infty (EA^{-1})^i\right) - \left( \sum_{i = 0}^\infty (EA^{-1})^{i+1}\right)\\
&= I - \lim_{i \rightarrow \infty} (EA^{-1})^i\\
&= I
\end{align*}
\end{proof}
\begin{rmk}
Recall that, because we rely on Lemma \ref{lem:lim0}, the meaning of this theorem is governed by Definition \ref{def:mat_conv}. That is what the equals sign really means. 
\end{rmk}
This fact makes sense. The spectral norm of $A^{-1}$ is the inverse of the smallest eigenvalue of $AA^*$, and it tells us how much $A$ is capable of reducing the euclidean norm of a vector it multiplies. So if that value is very large, $A$ is very nearly not invertible. In that case, it makes sense that adding a relatively small error matrix might change the inverse significantly.

We can use this result to determine how different the magnitude of a space similar to the one from Example \ref{ex:4pt} differed from that of an actual product space. Given matrices $A$ and $E$ which satisfy Theorem \ref{thm:approx}, we know that 
\[
(A - E)^{-1} - A^{-1} = A^{-1}\sum_{i = 1}^\infty (EA^{-1})^i.
\]
This means that we can let $\Phi = A^{-1}\sum_{i = 1}^\infty (EA^{-1})^i$ be the ``error term.'' If we can bound the sum of the entries of the matrix $\Phi$, that is a bound on the difference between the magnitude of a space whose similarity matrix is $A$, and one whose similarity matrix is $A + E$. Of course, this approximation scheme only works if both spaces admit M\"obius inversion.

Let $A$ be a generalized metric space which admits M\"obius inversion. Suppose we know $\chi(A)$, the magnitude of $A$, and we would like to know how much adding $E$, a perturbation matrix, to $\zeta_A$, will change the magnitude. Suppose that $E$ satisfies the condition of Theorem \ref{thm:approx}.

The magnitude of $A$ is equal to $||\mu_A||_1 = ||\zeta_A^{-1}||_1$, where $||\cdot||_1$ is the $L_1$ matrix norm.

>>>>

In fact, we observe exactly that behavior. Figure \ref{fig:prod_mag} shows the magnitudes of the three spaces as we vary $t$. At every value of $t$, $\chi_t(M)$ is almost exactly the product of the magnitudes of the other two spaces at that $t$ value.
\begin{figure}[H]
\[
\makebox[\textwidth][c]{\includegraphics[width = 0.9\paperwidth]{../Figures/prod_mag.png}}
\]
\caption{The magnitude of $M$ alongside the magnitudes of the factor spaces against $\log(t)$.}
\label{fig:prod_mag}
\end{figure}
\end{examp}

\section{Approximations and Error Bounding}
In Example \ref{ex:4pt}, $M$ was \emph{almost} a product space, so we expected its magnitude to be very nearly the product of the magnitudes of the two 2-point spaces. It would be useful to quantify how similar the magnitudes of two spaces are, given that we know something about how similar those spaces are. To do that, we need some simple tools from linear algebra involving vector and matrix norms. Note that these norms are not directly related to the norms we defined on monoidal categories - both are non-negative real-valued functions, but they operate on different structures.

There are many norms on matrices. Formally, a matrix norm is simply a function from matrices to the non-negative real numbers which satisfies certain properties.

@@

\item
\[
||A + B|| \leq ||A|| + ||B||.
\]
\item 
\[
||A|| = 0 \iff A = 0.
\]
\end{enumerate}
\end{lemma}
\begin{proof}
\begin{description}
\item[Spectral Norm:]
\begin{enumerate}
\item By the definition of the spectral norm, for any $v \in \Rr^n$,
\begin{align*}
|ABv| &\leq |A(||B||_s \times |v|)|\\
&\leq (||A||_s \times ||B||_s)|v|.
\end{align*}
It follows that 
 \[
||AB|| \leq ||A||_s\times ||B||.
\]
\item For any $v \in \Rr^n$,
\begin{align*}
|(A + B)v| &= |Av + Bv|\\
 &\leq |Av| + |Bv|\\
 &\leq ||A||_s + ||B||_s)v.
\end{align*}
\item It is clear that if $A=0$ then $||A||_s=0$. Now suppose that $A$ is not zero. Then the column space of $A$ has nonzero dimension, so there exists a $v$ such that $Av \neq 0$. So $||A||_s \geq \frac{|Av|}{|v|} > 0$.
\end{enumerate}
\item[$L_1$ Norm:]
\begin{enumerate}
\item We have
\begin{align*}
|AB_ij| &= \left| \sum_{k = 1}^nA_{ik}B{kj}\right|\\
&\leq \sum_{k = 1}^n|A_{ik}||B{kj}|. \text{ Therefore}\\
||AB||_1 &\leq \sum_{i = 1}^n\sum_{j=1}^n\sum_{k=1}^n|A_{ik}||B{kj}|\\
&\leq \sum_{i = 1}^n\sum_{j=1}^n\sum_{k=1}^n\sum_{l=1}^n|A_{ik}||B_{lj}|\\
&= ||A||_1||B||.
\end{align*}
\item For each $i,j$, $|(A + B)_{ij}| = |A_{ij} + B_{ij}| \leq |A_{ij}| + |B_{ij}|$. The result follows.
\item If $A=0$ then $||A||_1=0$. On the other hand, if any entry of $A$ is nonzero, then $||A||_1 > 0$ by definition.
\end{enumerate}
\end{description}
\end{proof}
Again, there are many other matrix norms. For our purposes, when a theorem uses an arbitrary matrix norm, we will assume that it satisfies Lemma \ref{lem:norm_prop}.

Now that we have some basic technology for working with matrix norms, we would like to show some results about limits and series convergence of matrices. To do that, we need to establish some notation. 
\begin{defn}\label{def:mat_conv}
Let $A_i$ be an infinite sequence of $m \times n$ matrices, and let $||\cdot||$ be some matrix norm (the spectral norm, the $L_1$ norm, or another matrix norm). Suppose we have an $m \times n$ matrix $L$ such that 
\[
\lim_{i \rightarrow \infty} ||A_i - L|| = 0.
\]
Then we say that 
\[
\lim_{i \rightarrow \infty} A_i = L.
\]
\end{defn}
It turns out that if the matrices involved are finite-dimensional, the choice of norm does not matter - a sequence will either always converge or always diverge. So there is no need to specify the norm when stating a fact about matrix convergence.

@@

\begin{lemma}\label{lem:lim0}
Let $E$ be a real-valued $n \times n$ matrix.  Suppose that for some matrix norm $||\cdot||$, $||E|| < 1$. Then:
\[
\lim_{i \rightarrow \infty}E^i = 0.
\]
\end{lemma}
\begin{proof}
Recall that we are supposing that any norm we use satisfies Lemma \ref{lem:norm_prop}.

Let $||E|| < 1$. Let $\vec{0}$ represent be the zero matrix. We now use Lemma \ref{lem:norm_prop}:
\begin{align*}
\lim_{i \rightarrow \infty}|| E^i - \vec{0}|| &\leq \lim_{i \rightarrow \infty}|| E^i|| - ||\vec{0}|| \\
&\leq \lim_{i \rightarrow \infty}|| E||^i\\
&= 0.
\end{align*}
Then, by Definition \ref{def:mat_conv},
\[
\lim_{i \rightarrow \infty} E^i = 0.
\]
\end{proof}

This allows us to prove a useful fact which allows us to approximate the inverses of certain matrices.
\begin{thm}\label{thm:approx}
Let $A$ be an invertible $n \times n$ matrix, let $||\cdot||$ be some matrix norm, and let $E$ be an $n \times n$ matrix such that $||E|| < \frac 1{||A^{-1}||}$. Then $(A-E)$ is invertible, and
\begin{align*}
(A - E)^{-1} &= A^{-1} + A^{-1}EA^{-1} + A^{-1}EA^{-1}EA^{-1} + \cdots\\
&= A^{-1} + A^{-1}\sum_{i = 1}^\infty (EA^{-1})^i.
\end{align*}
\end{thm}
\begin{proof}
We apply Lemma \ref{lem:lim0}
\begin{align*}
(A-E)  A^{-1}\sum_{i = 0}^\infty (EA^{-1})^i
&= \left(\sum_{i = 0}^\infty (EA^{-1})^i\right) - \left( \sum_{i = 0}^\infty (EA^{-1})^{i+1}\right)\\
&= I - \lim_{i \rightarrow \infty} (EA^{-1})^i\\
&= I
\end{align*}
\end{proof}
\begin{rmk}
Recall that, because we rely on Lemma \ref{lem:lim0}, the meaning of this theorem is governed by Definition \ref{def:mat_conv}. That is what the equals sign really means. 
\end{rmk}
This fact makes sense, in particular if we consider the spectral norm. $||A^{-1}||_s$ is the inverse of the smallest eigenvalue of $AA^*$, and it tells us how much $A$ is capable of reducing the euclidean norm of a vector it multiplies. So if that value is very large, $A$ is very nearly not invertible. In that case, it makes sense that adding a relatively small error matrix might change the inverse significantly.

Next, we use Theorem \ref{thm:approx} to find a closed-form bound on how much a perturbation matrix changes the magnitude of a generalized metric space.

Let $A$ be a generalized metric space which admits M\"obius inversion. Suppose we know $\chi(A)$, the magnitude of $A$, and we would like to know how much adding $E$, a perturbation matrix, to $\zeta_A$, will change the magnitude. Suppose that $||EA^{-1}|| < 1$ for some matrix norm, so we can use Theorem \ref{thm:approx}.

The magnitude of $A$ is equal to the sum of the entries of $\zeta_A^{-1}$. So the $L_1$ norm of the difference between $\zeta_A$ and $\zeta_A + E$ will be an upper bound on how much adding $E$ changes the magnitude of $A$. We can repeatedly use Lemma \ref{lem:norm_prop} to isolate the error term:
\begin{align}
\left|\left|\zeta_A^{-1} - \zeta_A^{-1} - \zeta_A^{-1}\sum_{i = 1}^\infty (EA^{-1})^i\right|\right|_1 \nonumber 
&= \left|\left|\zeta_A^{-1}\sum_{i = 1}^\infty (EA^{-1})^i\right|\right|_1 \nonumber \\
&\leq ||\zeta_A^{-1}||_1\sum_{i = 1}^\infty ||(EA^{-1})^i||_1. \label{eq:series}
\end{align}

Suppose that $||EA^{-1}||_1 <1$. Then we can convert \eqref{eq:series} to a closed form. 
\begin{align*}
||\zeta_A^{-1}||_1\sum_{i = 1}^\infty ||EA^{-1}||_1^i &\leq
||\zeta_A^{-1}||_1 \sum_{i = 1}^\infty||(\zeta_A + E)^{-1}||_1^i\\
 &\leq ||\zeta_A^{-1}||\frac{||EA^{-1}||_1}{1-||EA^{-1}||_1}.
\end{align*}

However, as we mentioned earlier, all matrix norms are equivalent when it comes to convergence. So if $||EA^{-1}||_1 \geq 1$ but, for example, $||EA^{-1}||_s <1$, the above would no longer work, but we would still expect \eqref{eq:series} to converge. This might be useful, because the spectral norm is often smaller than the $L_1$ norm. To make this work, we will need some way of replacing the $L_2$ norm with a spectral norm. It turns out we can do that:
\begin{lemma}\label{lem:norm_switch}
Let $A$ be an $n \times n$ matrix. Then
\[
||A||_1 \leq n^2||A||_s.
\]
\end{lemma}
\begin{proof}
Let $m$ be the absolute value of the largest element of $A$. Suppose that for some $i,j, |A_{ij}| > ||A||_s$. Let $e_j$ be the column vector with a one in index $j$ and zeros everywhere else. Then $Ae_j$ will have $A_{ij}$ in its $i$'th index, so $|Ae_j| \geq |A_{ij}| >||A||_s$. Because $|e_j|=1$, this is a contradiction. So $m \leq ||A||_1$. $||A||_1 \leq n^2m$, so $||A||_1 \leq n^2||A||_s$.
\end{proof}

This means that we can rewrite \eqref{eq:series} in the following way:
\begin{align*}
||\zeta_A^{-1}||_1\sum_{i = 1}^\infty ||EA^{-1}||_1^i &\leq n^2||\zeta_A^{-1}||_1\sum_{i = 1}^\infty ||(EA^{-1})^i||_s^i\\
&\leq n^2||\zeta_A^{-1}||_1\sum_{i = 1}^\infty ||(EA^{-1})||_s^i\\
&\leq n^2||\zeta_A^{-1}||\frac{||EA^{-1}||_s}{1-||EA^{-1}||_s}.
\end{align*}

We have (possibly) reduced the value of the norm, at the price of a factor of $n^2$. We can exchange the $L_1$ norm for other matrix norms in a similar fashion, using different constants.
